Effects of COVID-19 related social media use on well-being

### 2023-08-08 manuscript.Rmd
- shortened ms

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021].
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022].
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children [@kleinDarklySoothingCompulsion2021].
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question
[e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021],
<!-- [@bendauAssociationsCOVID19Related2021; @bradleyStressMoodSmartphone2021; @choiMediatedCommunicationMatters2021; @dornemannHowGoodBad2021; @edenMediaCopingCOVID192020; @guazziniSecondWaveAnalysis2022; @latikkaLonelinessPsychologicalDistress2022; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020; @sewallObjectivelyMeasuredDigital2021; @stainbackCOVID1924News2020; @yuePassiveSocialMedia2022] -->
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time returning to prior levels [@sheldonStabilityHappinessTheories2014].
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
So although well-being can be affected by external events and factors, this does not happen easily.
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.
<!-- For example, using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022]. -->
On the other hand, although mood management theory suggests that effects should not be overly negative, it could be that although short-term effects are positive, long-term effects are negative. 
<!-- it is still possible that people misjudge media effects. -->
<!-- Users are often overly optimistic [@metzgerComparativeOptimismPrivacy2017], focusing on short term benefits while ignoring long-term costs.  -->
Precisely because social media have so many positive consequences in the short run, this might cause problem in the long run---for example, because social media use displaces other potentially more meaningful and/or relaxing activities [@hallSocialMediaUse2022].
<!-- In other words, social media might not problematic because they are inherently bad, but rather because they are too good. -->
As with many other things, there can be too much of a good thing.
It is therefore often asked whether social media can become addictive, and users sometimes express this fear themselves [@yangCanWatchingOnline2021].
<!-- However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020]. -->
<!-- In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media use represent a new regular behavior that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018]. -->
<!-- In conclusion, there are several plausible theoretical mechanisms as to why effects might be negative or positive, explaining why on average effects might not be particularly negative or positive. -->

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
<!-- , outlining mechanisms of potential positive effects.  -->
<!-- At the same time, there are also several plausible reasons as to why COVID-19 related social media use is beneficial for well-being. -->
<!-- Access to information and resources:  -->
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
<!-- It allows individuals to stay informed about the latest updates, guidelines, and recommendations from reputable health organizations and experts.  -->
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
<!-- Educational opportunities:  -->
Social media can be a valuable _educational tool_, with various organizations and experts sharing informative content about COVID-19 [@worldhealthorganizationCoronavirusDiseaseCOVID192023]. 
<!-- This can include explanatory videos, infographics, and articles that help individuals understand the virus, its transmission, prevention strategies, and vaccination information.  -->
Access to educational resources on social media can empower individuals to make informed decisions regarding their health.
<!-- Community support and solidarity:  -->
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
<!-- Sharing experiences, offering support, and expressing solidarity can foster a sense of community and reduce feelings of isolation.  -->
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
<!-- Promoting mental health and well-being:  -->
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic [@twitterTalkingMentalHealthAwarenessTwitter2020]. 
<!-- They provide guidance on coping mechanisms, stress reduction techniques, and self-care practices.  -->
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
<!-- Encouraging positive behaviors:  -->
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@atheyDigitalPublicHealth2023; @huntSocialMediaBased2022]. 
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.
<!-- Amplifying diverse voices and perspectives:  -->
<!-- Social media platforms provide a space for diverse voices and perspectives to be heard [@brunsAreFilterBubbles2019].  -->
<!-- This allows individuals to access a range of viewpoints, experiences, and cultural practices related to COVID-19.  -->
<!-- Exposure to diverse perspectives might promote empathy, understanding, and a more comprehensive understanding of the pandemic's impact on different communities. -->

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
<!-- Misinformation and rumors:  -->
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
<!-- Overwhelm and anxiety:  -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022].
<!-- Confirmation bias and echo chambers:  -->
Social media algorithms are designed to show users content that aligns with their interests and beliefs, potentially contributing to _echo chambers_ [@cinelliEchoChamberEffect2021]. 
<!-- This might lead to confirmation bias, where individuals are more likely to be exposed to information that confirms their preexisting beliefs.  -->
In the context of COVID-19, this might reinforce false or misleading information and prevent individuals from considering alternative perspectives.
<!-- Cyberbullying and harassment:  -->
Social media platforms are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_ [@giumettiCyberbullyingSocialMedia2022]. 
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts. 
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
<!-- Social comparison and fear of missing out (FOMO):  -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].
<!-- Reduced face-to-face interaction:  -->
<!-- Spending excessive time on social media to learn about COVID-19 can lead to a decrease in face-to-face interactions and real-life connections [but see @hallSocialMediaUse2022].  -->
<!-- This might impact mental health, as most humans have an innate need for social support and physical presence, which online interactions might not fully replicate. -->

@@

<!-- As outline above,   -->
There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@kleinDarklySoothingCompulsion2021; @stainbackCOVID1924News2020; @yuePassiveSocialMedia2022]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, engaging more on social media was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
A couple of studies reported mostly neutral effects of social media use on well-being indicators [@bradleyStressMoodSmartphone2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

<!-- The hypothesis explicitly states a causal effect. -->
<!-- In non-experimental studies, longitudinal designs can help investigate causality. -->
When using longitudinal designs to analyze causality, it is important to (a) focus on within-person effects [@hamakerWhyResearchersShould2014]; to (b) control for confounders [@rohrerTheseAreNot2021]; and to (c) test a plausible interval between measures [@dormannOptimalTimeLags2015].
<!-- To illustrate, consider the following example. -->
<!-- Imagine that a person suddenly starts using social media much more than usual, and then after some time  becomes less satisfied with their life. -->
<!-- Eventually, use and life satisfaction return to prior levels. -->
<!-- If this happens to several people at the same time, in a longitudinal study we could then observe a significant effect of social media use on life satisfaction. -->
<!-- However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which caused large parts of the working population to loose their jobs. -->
<!-- Hence, the causal effect reported above was confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to go down. -->
<!-- Thus, only when controlling for _all_ relevant confounders, can we correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018]. -->
<!-- Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible. -->
First, in non-experimental designs it makes much sense to analyze causal effects from an internal, within-person perspective [@hamakerWhyResearchersShould2014].
If a specific person changes their media diet, we need to measure how this behavior affects their well-being.
Between-person comparisons from longitudinal data cannot provide such insights [@lucasWhyCrossLaggedPanel2022].
To test the hypothesis, I thus consider only the within-person effects.

Second, to identify confounders we should control for variables that affect both media use and well-being, which helps isolate the actual effect [@rohrerThinkingClearlyCorrelations2018].
Because we are adopting a within-person perspective, we need to implement _time-varying_ confounders [@rohrerTheseAreNot2021]. 
And because we are determining the _overall_ causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
<!-- Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider] -->
<!-- However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly. -->
<!-- The aim should therefore be to collect as many varying and non-varying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached. -->
<!-- [^collider]: In addition, there also exist colliders, which I don't discuss here and which complicate the issue even further [@rohrerThinkingClearlyCorrelations2018]. -->
<!-- We can also control for variables that affect only social media use or well-being.  -->
<!-- However, in doing so not much is gained, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021]. -->
In this study, I hence preregistered to control for the following variables, which either have already been shown or are likely to affect both social media use and well-being, and which also are not mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, residency Vienna, text-based news consumption, video-based news consumption, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, disposition to take risks, and locus of control [@egerStatisticalMetaanalysisWellbeing2015; @wardOriginsHappinessEvidence2016].
<!-- [^explore] -->
<!-- I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent. -->

@@

Life satisfaction was measured with the item "All things considered, how satisfied are you with your life as a whole nowadays?", which comes from the European Social Survey [@europeansocialsurveyESS9Edition20182021].
The response options ranged from 0 (_extremely dissatisfied_) to 10 (_extremely satisfied_).

@@

These specific observations notwithstanding, several general trends can be observed.
First, overall the results do suggest that effects of COVID-19 related social media use on well-being tend to take place in the negative as opposed to the positive spectrum.
Although very small, five statistically significant negative results of COVID-19 related social media use on well-being were found.
Only one positive effect emerged.
Also note that in the analyses several control variables were included, ruling out plausible alternative explanations for the negative results.
For example, it was controlled for as to whether or not participants contracted a COVID-19 infection during a specific wave.
Hence, we can rule out the alternative explanation that having an infection was the root cause of increased communication and reduced well-being.

Second, six significant outcomes emerged for positive or negative affect, but only one for life satisfaction. 
Life satisfaction is more stable and not that easily affected by any type or channel of social media communication.
The more fluctuating positive and negative affect, however, were affected (albeit only slightly).
Liking, sharing, and posting COVID-19 related content, and spending more time on Twitter and YouTube to browse COVID-19 related content, all slightly negatively influenced affect. 
This is aligned with prior findings which showed that social media use can trigger negative affect, but that it is less likely to determine life satisfaction [@huangTimeSpentSocial2017].
<!-- Writing posts implies more substantial engagement, and perhaps also negative exchanges with other users, which could explain the results. -->
<!-- For example, people who wrote more COVID-19 related posts than usual reported slightly lower levels of life satisfaction than usual. -->
<!-- Similarly, people who wrote more COVID-19 related posts than usual also reported slightly more negative affect. -->
<!-- As a potential explanation, when writing posts and comments on social media people explicitly and more deeply engage with COVID-19 related content. -->
Conversations about COVID-19 on social media are often extreme, negative, or aggressive [@fanStigmatizationSocialMedia2020].
More deeply engaging with this type of content could negatively affect active authors.
The hypothesis that tonality could explain the negative effects is especially supported by the observation that spending more time on Twitter and YouTube than usual increased negative affect.
Communication on both channels is often found to be negative and impolite [e.g., @muellerTwitterMadeMe2022], also when compared to other SNSs [@halpernSocialMediaCatalyst2013].
Consuming more negative and misleading information could hence explain the (slightly) increased levels of negative affect.

>>>>

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high [@statistaAverageDailyTime2021].
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged [@sharmaDarkEndTunnel2022].
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children.
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question
[e.g., @bendauAssociationsCOVID19Related2021; @edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021],
<!-- [@bendauAssociationsCOVID19Related2021; @bradleyStressMoodSmartphone2021; @choiMediatedCommunicationMatters2021; @dornemannHowGoodBad2021; @edenMediaCopingCOVID192020; @guazziniSecondWaveAnalysis2022; @latikkaLonelinessPsychologicalDistress2022; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020; @sewallObjectivelyMeasuredDigital2021; @stainbackCOVID1924News2020; @yuePassiveSocialMedia2022] -->
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are short-term, with well-being after some time returning to prior levels [@dienerAdvancesOpenQuestions2018].
Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
So although well-being can be affected by external events and factors, this does not happen easily.
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory [@zillmannMoodManagementCommunication1988], using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988].
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood [@marcianoDynamicsAdolescentsSmartphone2022]. 
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.
<!-- For example, using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022]. -->
<!-- INCLUDE IF SPACE -->
<!-- On the other hand, although mood management theory suggests that effects should not be overly negative, it could be that although short-term effects are positive, long-term effects are negative.  -->
<!-- Precisely because social media have so many positive consequences in the short run, this might cause problem in the long run---for example, because social media use displaces other potentially more meaningful and/or relaxing activities [@hallSocialMediaUse2022]. -->
<!-- As with many other things, there can be too much of a good thing. -->
<!-- It is therefore often asked whether social media can become addictive, with users sometimes expressing this fear themselves. -->
<!-- END -->
<!-- it is still possible that people misjudge media effects. -->
<!-- Users are often overly optimistic [@metzgerComparativeOptimismPrivacy2017], focusing on short term benefits while ignoring long-term costs.  -->
<!-- In other words, social media might not problematic because they are inherently bad, but rather because they are too good. -->
<!-- However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020]. -->
<!-- In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media use represent a new regular behavior that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018]. -->
<!-- In conclusion, there are several plausible theoretical mechanisms as to why effects might be negative or positive, explaining why on average effects might not be particularly negative or positive. -->

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
<!-- , outlining mechanisms of potential positive effects.  -->
<!-- At the same time, there are also several plausible reasons as to why COVID-19 related social media use is beneficial for well-being. -->
<!-- Access to information and resources:  -->
Despite incorrect information, social media provide a vast platform for disseminating _accurate and timely information_ about COVID-19 [@johnhopkinsuniversityCOVID19Map2023]. 
<!-- It allows individuals to stay informed about the latest updates, guidelines, and recommendations from reputable health organizations and experts.  -->
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
<!-- Educational opportunities:  -->
<!-- Social media can be a valuable _educational tool_, with various organizations and experts sharing informative content about COVID-19 [@worldhealthorganizationCoronavirusDiseaseCOVID192023].  -->
<!-- <!-- This can include explanatory videos, infographics, and articles that help individuals understand the virus, its transmission, prevention strategies, and vaccination information.  -->
<!-- Access to educational resources on social media can empower individuals to make informed decisions regarding their health. -->
<!-- Community support and solidarity:  -->
Social media platforms enable individuals to _connect with others_ who are experiencing similar challenges during the pandemic [@guazziniSecondWaveAnalysis2022]. 
<!-- Sharing experiences, offering support, and expressing solidarity can foster a sense of community and reduce feelings of isolation.  -->
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
<!-- Promoting mental health and well-being:  -->
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for _maintaining mental well-being_ during the pandemic [@twitterTalkingMentalHealthAwarenessTwitter2020]. 
<!-- They provide guidance on coping mechanisms, stress reduction techniques, and self-care practices.  -->
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
<!-- Encouraging positive behaviors:  -->
Social media campaigns and initiatives can promote _positive COVID-19 behaviors_, such as mask-wearing, physical distancing, hand hygiene, and vaccination [@atheyDigitalPublicHealth2023; @huntSocialMediaBased2022]. 
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.
<!-- Amplifying diverse voices and perspectives:  -->
<!-- Social media platforms provide a space for diverse voices and perspectives to be heard [@brunsAreFilterBubbles2019].  -->
<!-- This allows individuals to access a range of viewpoints, experiences, and cultural practices related to COVID-19.  -->
<!-- Exposure to diverse perspectives might promote empathy, understanding, and a more comprehensive understanding of the pandemic's impact on different communities. -->

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
<!-- Misinformation and rumors:  -->
Social media platforms can easily spread _false or misleading information_ about COVID-19 [@liYouTubeSourceInformation2020]. 
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
<!-- Overwhelm and anxiety:  -->
Constant exposure to COVID-19-related content on social media can lead to _information overload_ and contribute to heightened anxiety levels [@fanInformationOverloadWellbeing2021]. 
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic [@sharmaDarkEndTunnel2022].
<!-- Confirmation bias and echo chambers:  -->
<!-- Social media algorithms are designed to show users content that aligns with their interests and beliefs, potentially contributing to _echo chambers_ [@cinelliEchoChamberEffect2021].  -->
<!-- <!-- This might lead to confirmation bias, where individuals are more likely to be exposed to information that confirms their preexisting beliefs.  --> -->
<!-- In the context of COVID-19, this might reinforce false or misleading information and prevent individuals from considering alternative perspectives. -->
<!-- Cyberbullying and harassment:  -->
Social media platforms are known for fostering negativity, with users sometimes engaging in _cyberbullying and harassment_. 
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts. 
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
<!-- Social comparison and fear of missing out (FOMO):  -->
Social media often showcase the highlights and accomplishments of others, encouraging _social comparison_ [@przybylskiMotivationalEmotionalBehavioral2013]. 
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances [@sharmaDarkEndTunnel2022].
<!-- Reduced face-to-face interaction:  -->
<!-- Spending excessive time on social media to learn about COVID-19 can lead to a decrease in face-to-face interactions and real-life connections [but see @hallSocialMediaUse2022].  -->
<!-- This might impact mental health, as most humans have an innate need for social support and physical presence, which online interactions might not fully replicate. -->

@@

<!-- As outline above,   -->
There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison [@stainbackCOVID1924News2020; @yuePassiveSocialMedia2022]. 
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms [@bendauAssociationsCOVID19Related2021]. 
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes. 
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being [@guazziniSecondWaveAnalysis2022]. 
Additionally, engaging more on social media was associated with reduced feelings of loneliness [@latikkaLonelinessPsychologicalDistress2022].
Several studies reported mostly neutral effects of social media use on well-being indicators [@edenMediaCopingCOVID192020; @sewallObjectivelyMeasuredDigital2021].
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

<!-- The hypothesis explicitly states a causal effect. -->
<!-- In non-experimental studies, longitudinal designs can help investigate causality. -->
When using longitudinal designs to analyze causality, it is important to (a) focus on within-person effects [@hamakerWhyResearchersShould2014]; to (b) control for confounders [@rohrerTheseAreNot2023]; and to (c) test a plausible interval between measures [@dormannOptimalTimeLags2015].
<!-- To illustrate, consider the following example. -->
<!-- Imagine that a person suddenly starts using social media much more than usual, and then after some time  becomes less satisfied with their life. -->
<!-- Eventually, use and life satisfaction return to prior levels. -->
<!-- If this happens to several people at the same time, in a longitudinal study we could then observe a significant effect of social media use on life satisfaction. -->
<!-- However, it could also be the case that during the study there was a major exogenous event (say, a pandemic), which caused large parts of the working population to loose their jobs. -->
<!-- Hence, the causal effect reported above was confounded, because in reality it was the pandemic that caused both social media use to rise and life satisfaction to go down. -->
<!-- Thus, only when controlling for _all_ relevant confounders, can we correctly estimate causality without bias [@rohrerThinkingClearlyCorrelations2018]. -->
<!-- Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible. -->
First, in non-experimental designs it makes much sense to analyze causal effects from an internal, within-person perspective [@hamakerWhyResearchersShould2014].
If a specific person changes their media diet, we need to measure how this behavior affects their well-being.
Between-person comparisons from longitudinal data cannot provide such insights [@hamakerWhyResearchersShould2014].
To test the hypothesis, I thus consider only the within-person effects.

Second, to identify confounders we should control for variables that affect both media use and well-being, which helps isolate the actual effect [@rohrerThinkingClearlyCorrelations2018].
Because we are adopting a within-person perspective, we need to implement _time-varying_ confounders [@rohrerTheseAreNot2023]. 
And because we are determining the _overall_ causal effect, we need to make sure _not_ to control for mediating variables [@rohrerThinkingClearlyCorrelations2018], for doing so would bias our assessment of the causal effect.
<!-- Complicating matters further, it is often unclear if a variable is a mediator or a confounder.[^collider] -->
<!-- However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly. -->
<!-- The aim should therefore be to collect as many varying and non-varying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached. -->
<!-- [^collider]: In addition, there also exist colliders, which I don't discuss here and which complicate the issue even further [@rohrerThinkingClearlyCorrelations2018]. -->
<!-- We can also control for variables that affect only social media use or well-being.  -->
<!-- However, in doing so not much is gained, because the effects of social media use would remain virtually the same [@klinePrinciplesPracticeStructural2016; but see @mcelreathYesterdayClass2021]. -->
In this study, I hence preregistered to control for the following variables, which either have already been shown or are likely to affect both social media use and well-being, and which also are not mediators:
<!-- (I'll additionally include variables that likely affect only well-being, also to obtain a comparison benchmark for social media effects): -->
gender, age, education, Austria country of birth, Austria country of birth of parents, residency Vienna, text-based news consumption, video-based news consumption, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, disposition to take risks, and locus of control [@egerStatisticalMetaanalysisWellbeing2015].
<!-- [^explore] -->
<!-- I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent. -->

@@

Life satisfaction was measured with the item "All things considered, how satisfied are you with your life as a whole nowadays?", which comes from the European Social Survey.
The response options ranged from 0 (_extremely dissatisfied_) to 10 (_extremely satisfied_).

@@

These specific observations notwithstanding, several general trends can be observed.
First, overall the results do suggest that effects of COVID-19 related social media use on well-being tend to take place in the negative as opposed to the positive spectrum.
Although very small, five statistically significant negative results of COVID-19 related social media use on well-being were found.
Only one positive effect emerged.
<!-- Also note that in the analyses several control variables were included, ruling out plausible alternative explanations for the negative results. -->
<!-- For example, it was controlled for as to whether or not participants contracted a COVID-19 infection during a specific wave. -->
<!-- Hence, we can rule out the alternative explanation that having an infection was the root cause of increased communication and reduced well-being. -->

Second, six significant outcomes emerged for positive or negative affect, but only one for life satisfaction. 
Life satisfaction is more stable and not that easily affected by any type or channel of social media communication.
The more fluctuating positive and negative affect, however, were affected (albeit only slightly).
Liking, sharing, and posting COVID-19 related content, and spending more time on Twitter and YouTube to browse COVID-19 related content, all slightly negatively influenced affect. 
This is aligned with prior findings which showed that social media use is associated with increased negative affect but not with life satisfaction [@meierComputermediatedCommunicationSocial2020a].
<!-- Writing posts implies more substantial engagement, and perhaps also negative exchanges with other users, which could explain the results. -->
<!-- For example, people who wrote more COVID-19 related posts than usual reported slightly lower levels of life satisfaction than usual. -->
<!-- Similarly, people who wrote more COVID-19 related posts than usual also reported slightly more negative affect. -->
<!-- As a potential explanation, when writing posts and comments on social media people explicitly and more deeply engage with COVID-19 related content. -->
Conversations about COVID-19 on social media are often extreme, negative, or aggressive [@fanStigmatizationSocialMedia2020].
More deeply engaging with this type of content could negatively affect active authors.
The hypothesis that tonality could explain the negative effects is especially supported by the observation that spending more time on Twitter and YouTube than usual increased negative affect.
Communication on both channels is often found to be negative and impolite [e.g., @muellerTwitterMadeMe2022].
Consuming more negative and misleading information could hence explain the (slightly) increased levels of negative affect.

Third, the results show that it makes sense to analyze different communication types and communication channels.
Reading slightly reduced positive affect, while liking, sharing, and posting slightly increased negative affect.
Interestingly, posting COVID-19 related comment slightly increased negative affect, while at the same time it also slightly increased positive affect.
Posting content is often met with strong reaction, both positive by means of likes and negative by means of critical comments. 
Overall, though, posting led to slightly reduced levels of life satisfaction.
In conclusion, whereas it was often stated that passive use is bad and active use good [@verduynPassiveFacebookUsage2015], this pattern was only partially found here.
<!-- Instead, all three types were negative. -->
The results are aligned with the findings from @valkenburgAssociationsActivePassive2022, who could not confirm that active use is good and that passive use is bad.
Focusing on communication channels, Twitter and YouTube seem to be more negative, while Instagram, WhatsApp, and Facebook were neutral.
But, again, all of these effects are very small.
<!-- Future research might elaborate on these specific relations to probe their stability and relevance. -->
